---
title: "News Consumption and Federal Election Interest in 2015 CES"
author: "Alen Mitrovski"
date: 2020-09-23T21:13:14-05:00
bibliography: "references.bib" 
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# Author: Alen Mitrovski

# Abstract

In this paper, I looked into the distributions of news consumption and election interest in the 2015 Canadian Election Study. Using R, I was able to graph these results and examine their distributive properties. I hypothesized that the distributions of news consumption and interest in the federal election will both be right skewed. As a result of examining the graphs, I concluded that the distribution of federal election interest was right skewed, whereas, the distribution of news consumption followed a normal distribution.

# Introduction

The 2015 Canadian federal election was held October 19, 2015. Liberal party leader, Justin Trudeau, ended the nine-year reign of Stephen Harper to become the 23rd Canadian Prime Minister (Zurcher, 2015). The Canadian Election Study (CES) is a survey conducted to examine Canadians’ opinions regarding political parties, election campaigns and party leaders (Northrup, 2016). This study has been an integral part of Canadian democracy research since 1965 (Fournier, 2015). 

As per reports, voter turnout increased 7.2 percent from 2011 to 2015, marking the largest rise in turnout since 1900 (Voter Turnout). I am interested in exploring whether surveys of Canadians’ political interest correlate with this increase in voter turnout in the 2015 election. Additionally, I would like to follow up by exploring whether news consumption could be a factor for the distribution of interest in the 2015 Canadian federal election. 

In order to investigate these properties, I will graph two survey results from the 2015 CES. The first is the distribution of interest in the 2015 federal election. The second graphed result is the distribution of news consumption in the 2015 CES. I will be comparing their distributive properties to see if there is a correlation between news consumption and interest in the federal election.

I hypothesize that:
  
  1.	The distribution of interest in the 2015 federal election will represent a right-skewed distribution. The null hypothesis is that the distribution of interest in the 2015 federal election will not represent a right-skewed distribution. 
  2.	The distribution of news consumption in the 2015 CES will represent a right-skewed distribution. The null hypothesis is that the distribution of news consumption in the 2015 CES  will not represent a right-skewed distribution. 
  
# Graphs

```{r}
install.packages("tidyverse")

install.packages("devtools") 
devtools::install_github("hodgettsp/cesR")

library(tidyverse)
library(cesR)
library(labelled)

get_ces("ces2015_web")

ces2015_web <- to_factor(ces2015_web)
```


```{r}
p <- ces2015_web %>% 
  ggplot(aes(x = p_mediatim)) + 
  geom_bar()

d <- p + ggtitle("Distribution of News Consumption in 2015 CES") + xlab("Dialy Time Spent") + ylab("Individuals")

d <- d + theme_grey(base_size = 8)
plot(d)

```
  
In this graph, the distribution of news consumption in the 2015 CES follows a normal distribution. The largest portion of individuals spend 30-60 minutes watching, reading and listening to the news. In addition, over 3000 individuals chose not to answer this survey question.   
  
```{r}
a <- ces2015_web %>% 
  ggplot(aes(x = interest)) + 
  geom_bar()

b <- a + ggtitle("Distribution of Federal Election Interest in 2015 CES") + xlab("Level of Interest") + ylab("Individuals")

b <- b + theme_grey(base_size = 8)
plot(b)
```  
  
In this graph, we find that the distribution of federal election interest in the 2015 CES follows a right-skewed distribution. Approximately 3500 individuals reported interest levels between 7-10 out of 10. In addition, 500 individuals chose not to answer this survey question.   
  
# Discussion

Through the graphical evidence, we reject the null hypothesis that the distribution of interest in the 2015 federal election will not represent a right-skewed distribution. As Elections Canada has found, voting in the election and political interest are, in fact, related (Explaining the Turnout). This is further exemplified in the graph of the distribution of federal election interest. The right-skewed nature of this graph shows that the majority of individuals reported between 7-10 level of interest in the upcoming election. This high level of interest could be a reason as to why there was a reported 7.4 percent increase in voter turnout from 2011 to 2015. According to Elections Canada, 16.1 percent of individuals who did not vote in the election cited a lack of interest as the catalyst for their decision (Explaining the Turnout). Therefore, interest is a large contributing factor to voter turnout. 

We fail to reject the null hypothesis that the distribution of news consumption in the 2015 CES will not represent a right-skewed distribution. I had initially believed that the high levels of interest surrounding this 2015 election could be contributed to a high volume of daily news consumption. However, as seen in the two graphs above, the distributive properties of these two data sets do not line up. While news consumption follows a normal distribution, the interest levels in the election follow a right-skewed distribution. This may suggest that higher number of hours spent watching, reading or listening to the news is not correlated with higher levels of interest in the election. Curtis et al. had previously concluded that there is a positive correlation between age and political interest for Canadians (Curtis, 1976). However, thus far, there is not much research in the subject of election interest. 


# Weaknesses

The main limitations of this work include the following:

  1.	Considerable amount of “Not Answered” and “Refuse to Answer” results in both the Interest and News Consumption charts. Individuals seem unwilling to answer certain questions. 
  2.	There could be confounding factors that contribute noise to the data in the CES. For instance, the CES study included attention-based questions that told the subjects what answer to respond with. However, these questions were not answered with 100 percent coherence. Thus, it can be extrapolated that not all subjects were answering with complete attention and consistency to the study. 
  3.	There could be human error in writing the code in R to create the graphs used in this report. 
  
# Next Steps 

The future steps would be the following:

  1.	Reducing the number of potential response answers (Reference as to why this is beneficial). Some individuals might lose their patience reading through a thorough survey, such as the CES, and thus resort to simple “NA” answers. To mitigate for this, researchers can limit their potential answers to individual questions to under five possible answers. In addition, reducing the overall question count may increase the response rate to certain questions.  
  2.	It may be insightful to perform a follow-up multi-variable study looking into the various factors that led to the increase in voter turnout from 2011 to 2015. As a result, it will be possible to see how much federal election interest has affected this increase in voters during this period.  
  3.	Researchers should examine what factors persuade an individual to be more interested in a federal election. Unlike I initially hypothesize, news consumption does not seem to be an indicator of interest levels for Canadians. 
  4.	It would be particularly interesting to compare the distributions of interest levels in the 2015 Canadian federal election to previous years to see what sort of trend has been occurring over the years. Likewise, it would be interesting to break down the different forms of news consumption (ie. paper, electronic, etc.) and see how their consumption has been trending over the past couple decades as technology has advanced. 

# References

  1.	Zurcher, Anthony. “Trudeau Brings Liberals Back on Top.” BBC News, BBC, 20 Oct. 2015, www.bbc.com/news/world-us-canada-34566369.
  2.	Northrup, David. "The 2015 Canadian Election Study." Toronto, Ontario, Canada: The Institute for Social Research, York University (2016).
  3.	Fournier, Patrick, Fred Cutler, Stuart Soroka and Dietlind Stolle. 2015. The 2015 Canadian Election Study. [dataset]
  4.	Explaining the Turnout Decline in Canadian Federal Elections: A New Survey of Non-voters. Elections Canada. Accessed September 26, 2020. https://www.elections.ca/content.aspx?section=res&dir=rec/part/tud&document=interest&lang=e
  5.	Voter Turnout at Federal Elections and Referendums. Elections Canada. Accessed September 26, 2020. https://www.elections.ca/content.aspx?section=ele&dir=turn&document=index&lang=e
  6.	Curtis, James E., and Ronald D. Lambert. “Voting, Election Interest, and Age: National Findings for English and French Canadians.” Canadian Journal of Political Science, vol. 9, no. 2, 1976, pp. 293–307., doi:10.1017/s0008423900043730.
  7.	Paul A. Hodgetts and Rohan Alexander (2020). cesR: Access the CES Datasets a Little Easier.. R package version 0.1.0.
  8.	Hadley Wickham, Jim Hester and Winston Chang (2020). devtools: Tools to Make Developing R Packages Easier. https://devtools.r-lib.org/,https://github.com/r-lib/devtools.
  9.	R Core Team (2020). R: A language and environment for statistical computing. R, Foundation for Statistical Computing, Vienna, Austria. URL, https://www.R-project.org/.
  10.	JJ Allaire and Yihui Xie and Jonathan McPherson and Javier Luraschi and Kevin Ushey and Aron Atkins and Hadley Wickham and Joe Cheng and Winston Chang and Richard Iannone (2020). rmarkdown: Dynamic Documents for R. R package version 2.3. URL https://rmarkdown.rstudio.com.

